The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-19 paper.qmd
fixed choropleth legend

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---

@@

# load Libraries
library(tidyverse)
library(biscale)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
library(urbnmapr)
```
# Background

@@

Unlike an infographic, a statistical graphic is one of intent, and all enhancements made on a statistical graphic should be in pursuit of that goal (@Vanderplas2020).

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-ncol: 3

@@

Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarize the field. @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. Additionally, they noticed that the literature leaves@Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers. @uncertchap2022 summarized several cognitive effects that have repeatedly arised in uncertainty literature, however these effects were each discussed in isolation as a list of considerations an author might make. While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly state goal of uncertainty visualisation, scientific transparency. The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.

@@

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depicts the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."

@@

p1a
p1b
show_col(basecols, ncol=8, labels=FALSE)
```

>>>>

---
title: "The Noisy Work of Uncertainty Visualisation Research: A Review"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
fig-valign: bottom
editor_options: 
  chunk_output_type: console
---

@@

# load Libraries
library(tidyverse)
library(biscale)
library(RColorBrewer)
library(scales)
library(sf)
library(ggrepel)
library(urbnmapr)
```

{{< pagebreak >}}

# Background

@@

Visualization is an important step in exploratory data analysis and it is often utilised to **learn** what is important about a data set. Datasets such as Anscombe's quartet [@anscombe] or the Datasaurus Dozen [@datasaurpkg] highlight this power in visualisation. Additionally, visualisations allow for efficient and memorable communication. Even something as simple as sketching a distribution before recalling statistics or making predictions can greatly increase the accuracy of those measures [@Hullman2018; @Goldstein2014].

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-ibrekk
#| fig-cap: "A replication of the the uncertainty visualisations shown by @Ibrekk1987 in one of the earliest uncertainty visualisation experiments. This early experiment is a good example of many of the issues that are still common in uncertainty visualisation today. For example, the '95% confidence interval' is more accurately a '95% prediction interval'. Additionally graphics that depict different mathematical objects that are also different on their visual components are compared because of a percieved relation to uncertainty. So visualisations of the mean, PDF and CDF, are all discussed as though they all contain the relevant statistical information. The axis also have different scales, and visualisation methods that are now unpopular for displaying proportions, such as a pie chart, are used."
#| fig-subcap: 
#|   - "Picture 1"
#|   - "Picture 2"
#|   - "Picture 3"
#|   - "Picture 4"
#|   - "Picture 5"
#|   - "Picture 6"
#|   - "Picture 7"
#|   - "Picture 8"
#|   - "Picture 9"
#| layout-valign: bottom
#| layout-ncol: 3

@@

Reviews on uncertainty visualisation rarely offer tried and tested rules for effective uncertainty visualisation, but rather comment on the *difficulties* faced when trying to summarize the field. @Kinkeldey2014 found most experiments on the methods for uncertainty visualisation evaluation to be ad hoc, with no commonly agreed upon methodology or formalisation and no greater goal of describing general principals. @Hullman2016 commented on the difficulty in taking overarching themes from uncertainty visualisation, as several conflated issues make it unclear if subjects did poorly in an experiment because they misunderstood a visualisation, because the question was misinterpreted, or because they used a specific heuristic. @Spiegelhalter2017 commented that different plots are good for different things, and disagreed with the goal of identifying a universal "best" plot for all people and circumstances. @Griethe2006 was unable to find common themes, but instead listed the findings and opinions of a collection of papers. @uncertchap2022 summarized several cognitive effects that have repeatedly arised in uncertainty literature, however these effects were each discussed in isolation as a list of considerations an author might make. While these reviews are thorough in scope, none discuss how the existing literature contribute to the commonly state goal of uncertainty visualisation, scientific transparency. The problem faced by the literature is easily summarized with a famous quote by Henri Poincaré.

@@

{{< pagebreak >}}

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: fig-choropleth
#| fig-cap: "Two choropleth maps that depicts the counties of Iowa where each country coloured acording to a simulated average temperature. Both maps depict a spatial trend, where counties closer to the center of the map are hotter than counties on the edge of the map. This trend is only technically there in the low variance condition, i.e. if we performed a hypothesis test we would conclude the trend exists. The high variance condition displays a spatial trend that could simply be the result of noise, which means the plot is displaying a false conclusion."
#| fig-subcap: 
#|   - "High Variance Data"
#|   - "Low Variance Data"
#|   - ""
#| layout-ncol: 3
#| layout-valign: "bottom"

@@

show_pal1 <- function (colours, borders = NULL, cex_label = 1, 
    ncol = NULL) {
  # Set dimensions of palette
  n <- length(colours)
  ncol <- ncol %||% ceiling(sqrt(length(colours)))
  nrow <- ceiling(n/ncol)
  # make matrix with null values (if not full)
  colours <- c(colours, rep(NA, nrow * ncol - length(colours)))
  colours <- matrix(colours, ncol = ncol, byrow = TRUE)
  # set graphical parameters (?)
  old <- par(pty = "s", mar = c(0, 0, 0, 0))
  on.exit(par(old))
  size <- max(dim(colours))
  plot(c(0, size), c(0, -size), type = "n", xlab = "", ylab = "", 
      axes = FALSE)
  rect(col(colours) - 1, -row(colours) + 1, col(colours), -row(colours), 
       col = colours, border = borders)
  text(c(0,col(colours)), -c(1,row(colours))-0.15, 21:29, 
              cex = 1, col = "black")
  text(4, -1.55, "Temperature",cex = 1.5, col = "black")
}

p1a
p1b
show_pal1(basecols, ncol=8, borders=NA)

```
